---
title: "R | Selenium"
description: |
  Programmatically drive a web browser
author:
  - name: shikokuchuo
date: 05-03-2021
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
par(mar = c(0, 0, 0, 0))
plot(imager::load.image(photos), axes = FALSE)

```
<aside>Shikokuchuo</aside>

### Use case

Whenever you need to programmatically drive a web browser.

Most often:

- to scrape information behind a login screen
- when the http server does not return a simple html document

### Initial setup

Prerequisites: JRE or JDK installed on your system, Mozilla Firefox

1. Install the [RSelenium](https://github.com/ropensci/RSelenium) package from CRAN:

```{r rselenium, eval=FALSE}

install.packages("RSelenium")

```

2. Go to https://selenium-release.storage.googleapis.com/index.html

Download selenium-server-standalone-4.0.0-alpha-2.jar (or whatever is the latest 'selenium-server-standalone' file)

3. Go to https://github.com/mozilla/geckodriver

Download the latest Mozilla geckodriver release, and place in same directory as the jar file

### Running Selenium Webdriver

At the terminal, first cd to the directory where your two new files are saved, then run: 

```{r rseleniumt, eval=FALSE}

java -jar selenium-server-standalone-4.0.0-alpha-2.jar

```

The selenium server must be up and running before attempting to execute the R code below.

### RSelenium quickstart code

```{r rseleniumqs, eval=FALSE}

library(RSelenium)
library(keyring)
library(rvest)
library(magrittr)

# Start Selenium Session
remDr <- remoteDriver(
  remoteServerAddr = "localhost",
  port = 4444L,
  browserName = "firefox"
)

remDr$open()

# Navigate to login page
remDr$navigate("https://website.com/login")
Sys.sleep(5) # Give page time to load

# Find 'username' element and send 'saved_user' as input
webElem1 <- remDr$findElement(using = "xpath", "//input[@name = 'username']")
webElem1$sendKeysToElement(list(key_get("saved_user")))

# Find 'password' element and send 'saved_pass' and 'enter' keystroke as input
webElem2 <- remDr$findElement(using = "xpath", "//input[@name = 'password']")
webElem2$sendKeysToElement(list(key_get("saved_pass"), key = "enter"))
Sys.sleep(5) # Give page time to load

# Navigate to desired page and download source
remDr$navigate("https://website.com/somepage")
Sys.sleep(5) # Give page time to load
html <- remDr$getPageSource()[[1]] %>% read_html()

# Use further rvest commands to extract required data
# ...

# End Selenium Session
remDr$close()

```
<aside>
1. Customise the URLs as required.

2. Customise the xpath to locate the desired input fields as they are actually named on your site.

3. 'saved_user' and 'saved_pass' are values already stored using the [keyring](https://github.com/r-lib/keyring) package and retrieved here using the 'key_get' command. It is never a good idea to store plain text credentials in an R script.
</aside>

### Reference

Basic vignette: https://docs.ropensci.org/RSelenium/articles/basics.html

