---
title: "secretbase: The 40KB Hash Package"
description: |
  Streaming cryptographic hashes for R
author: shikokuchuo
date: 2026-02-04
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
      ________
     /\ sec   \
    /  \ ret   \
    \  /  base /
     \/_______/
```

At 40KB - smaller than most hex stickers - {secretbase} delivers streaming cryptographic hashes for R. It does one thing, and does it well: hash R objects and files without memory overhead.

R already has excellent packages for cryptographic hashing. The {digest} package by Dirk Eddelbuettel has served the community since 2003. The {openssl} package by Jeroen Ooms provides the full OpenSSL library. {secretbase} complements these with a minimal, dependency-free implementation built around streaming from the ground up.

### Zero-Copy Streaming

When you hash an R object using most approaches, the object must first be serialized to a raw vector, which is then hashed. For large objects, this means allocating memory for the entire serialized representation before hashing can begin.

{secretbase} eliminates this intermediate step. Using R's internal serialization callbacks, data flows directly from the serialization process into the hash computation:

```{r streaming}
library(secretbase)

# Hash a large object without materializing the serialized form
large_data <- data.frame(x = rnorm(1e6), y = rnorm(1e6))
sha3(large_data)
```

No memory spike. The data frame is never converted to a serialized representation in memory - serialization feeds directly into the hash context in 64KB chunks.

The same applies to files. Hash files larger than available RAM:

```{r file, eval=FALSE}
sha256(file = "huge_dataset.parquet")
```

For reproducibility, hash values are consistent across platforms: always R serialization version 3, big-endian byte order, skipping headers. The same object produces identical hashes on Linux, macOS, and Windows.

### Powering Reproducible Pipelines

{secretbase} serves as infrastructure for the {targets} package, Will Landau's Make-like pipeline tool for data science workflows.

{targets} determines whether pipeline targets need rebuilding by comparing hash values. When code or data changes, {targets} recomputes the hash - if it matches the stored value, the target is skipped. This happens constantly during pipeline execution, so hashing performance directly impacts the time spent checking versus computing.

Streaming serialization matters here. A pipeline checking dozens of large data frames on each run would otherwise serialize each object to memory before hashing. With {secretbase}, memory footprint stays constant regardless of object size. Cross-platform consistency ensures pipelines produce identical results whether developed on a laptop or executed on a cluster.

### Hash Functions

{secretbase} provides SHA-3 (NIST 2015), SHA-256, SHAKE256 (extendable-output), SipHash-1-3 (fast keyed hashing), and Keccak (used by Ethereum):

```{r hashes}
sha3("secret base")

sha256("secret base")

# SHAKE256 can produce any output length - useful for deterministic seeds
shake256("my seed", bits = 32L, convert = NA)
```

### Encoding Functions

Beyond hashing, {secretbase} provides encoding utilities:

**Base64** - Standard base64 encoding and decoding:

```{r base64}
base64enc("hello world")

base64dec("aGVsbG8gd29ybGQ=")
```

**Base58 with checksum** - Base58 encoding with a 4-byte double-SHA256 checksum for error detection:

```{r base58}
base58enc("secret base")

# Checksum verification on decode
base58dec(base58enc("secret base"))
```

Both base64 and base58 encoding functions support arbitrary R objects through serialization:

```{r roundtrip}
# Round-trip any R object through base64
obj <- list(a = 1:10, b = "test")
identical(obj, base64dec(base64enc(obj), convert = NA))
```

**CBOR** - Concise Binary Object Representation (RFC 8949) is a binary data serialization format designed to be compact and efficient to parse. Think of it as a binary alternative to JSON - it represents the same data model (maps, arrays, strings, numbers) but in a more space-efficient binary encoding.

CBOR was developed by the IETF and has become a standard in several domains:

- **WebAuthn/FIDO2** - The passwordless authentication standard
- **COSE** - CBOR Object Signing and Encryption, used in security protocols
- **HTTP/3 and QUIC** - Used in various web protocol extensions

{secretbase} brings CBOR support to R:

```{r cbor}
# Round-trip any R object through CBOR
obj <- list(a = 1L, b = "hello", c = TRUE)
identical(obj, cbordec(cborenc(obj)))
```

### Implementation

{secretbase} is implemented in C, with hash algorithm implementations derived from:

- [Mbed TLS](https://www.trustedfirmware.org/projects/mbed-tls/) (SHA-3, SHA-256, Base64)
- [c-siphash](https://github.com/c-util/c-siphash) (SipHash)
- [libbase58](https://github.com/luke-jr/libbase58) (Base58)

The R interface is minimal: thin wrapper functions that pass directly to `.Call()` with no R-code processing overhead. The package has no dependencies beyond base R, keeping installation simple and avoiding potential conflicts.

### Links

- {secretbase} website: <https://shikokuchuo.net/secretbase/>
- {secretbase} on CRAN: <https://cran.r-project.org/package=secretbase>
- {targets} website: <https://docs.ropensci.org/targets/>

---

*40KB. No dependencies. One thing, done well.*
