---
title: "Reverting Git Commits"
description: |
  Procedure to roll back both local and remote changes
author:
  - name: shikokuchuo
date: 2021-08-12
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
par(mar = c(0, 0, 0, 0))
plot(imager::load.image(photos), axes = FALSE)

```
<aside>Shikokuchuo</aside>

### Reverting Local Git Commits

You have made a commit.

You discover a mistake or something you left out straight after the commit.

```{r, eval = FALSE}

git reset HEAD~

```

This is a soft reset. Your changes are preserved. The commit is removed from the record.

Make the additional changes you need. Add files. Commit.


### Reverting Commits Pushed to Remote (e.g. Github)

Copy your folder to a backup location.

The following is a hard reset, which rolls back to the previous commit. Changes since that commit will be lost. Force push it to the remote.

```{r, eval = FALSE}

git reset HEAD^ --hard
git push origin -f

```

Both local and remote should now be in sync at the previous commit. You may check with:

```{r, eval = FALSE}

git status

```

If you have Github Actions that are triggered by commits, they will be triggered again despite this being a roll-back. So go and stop those runs if necessary.

Next, if you have another branch such as 'gh-pages' that builds automatically on each commit, roll back that branch as well so it keeps in sync. As this branch has been building on the remote, do a git pull to ensure that your local copy is up to date first before resetting.

```{r, eval = FALSE}

git checkout gh-pages
git pull

git reset HEAD^ --hard
git push -f origin gh-pages

```

Check status. Switch back to 'main' branch (substitute whatever branch you were on).

```{r, eval = FALSE}

git status
git checkout main

```

Copy back the files with changes you made previously from your backup location.

Make the additional changes you need. Add files. Commit.
