---
title: "do.call / lapply"
description: |
  A distinctive coding style for R
author:
  - name: shikokuchuo
date: 05-28-2021
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
develop <- function(x) {
  plot(imager::load.image(x), axes = FALSE)
  paste0(openssl::sha256(file(x)))
}
par(mar = c(0, 0, 0, 0))
data.frame(sha256 = develop(photos))

```
<aside> Shikokuchuo </aside>

### Use case

The use of the do.call / lapply combination is a powerful way to leverage functional programming in R. In short, you write a function that performs some actions and apply it to a list of inputs, which can then be fed into a function that combines everything into a single object.

Let us take an example, where we would like to calculate the ichimoku clouds for a portfolio of stocks, but also preserve the volume data, all in one tidy object.

### Setup

We could set it up as per the below:

- tickers: a vector defining the stock symbols in our portfolio
- process: a function that generates a row in a data frame or matrix

```{r functional, echo=TRUE}
library(ichimoku)

tickers <- c("C", "MS", "JPM", "GS")

process <- function(x) {
  # Use the 'quantmod' package to download pricing data
  pxdata <- quantmod::getSymbols(x, from = "2020-04-15", to = "2021-05-27", auto.assign = FALSE)
  # Extract volume column
  volume <- pxdata[, grep("Volume", colnames(pxdata))]
  # Calculate the cloud by calling ichimoku() from the 'ichimoku' package
  cloud <- ichimoku(pxdata, ticker = x)
  # Return a list of ticker, ichimoku cloud object, volume data
  list(x, cloud, volume)
}

```
<aside> Note: the original pricing data is preserved within the ichimoku object.</aside> 

We now want to apply our function to each element of 'tickers' in turn, and then for the results to be combined.

### Loop

One way to achieve this would be to iterate over 'tickers' using a loop:

```{r loop, echo=TRUE}

# Define a list to contain the loop output, specifying the length in advance as good practice
out <- vector(mode = "list", length = length(tickers))

# Loop over each element in 'tickers' and save in pre-defined list
for(i in seq_along(tickers)) out[[i]] <- process(tickers[i])

# Create output matrix by calling rbind on each element of the list
portfolio <- do.call(rbind, out)

portfolio

```

This approach takes 3 lines of code.

Furthermore, 'i' and 'out' remain as leftover objects in the global environment.

Somewhat messy.

### do.call / lapply

Instead we can use a do.call / lapply combination to achieve the same result in one line:

```{r lapply, echo=TRUE}

portfolio <- do.call(rbind, lapply(tickers, process))

portfolio

```

There are also no intermediate objects generated that clutter the global environment.

To explain:

- First lapply applies to a list or list-like object ('tickers'), a function ('process'). lapply always returns a list.

- This can then be fed into do.call, which calls a function ('rbind') on a list of arguments (the output of lapply i.e. the lists returned by 'process').

The use of do.call / lapply provides for a far more succinct and distinctive coding style.

### Tidy data output

```{r portfolio, echo=TRUE}
portfolio
```

'portfolio' is a tidy matrix with a row for each ticker, and a column for each data type.

We can easily access any element of the matrix by specifying its index value, for example the ichimoku cloud for MS by [2,2]:

```{r output, echo=TRUE}
plot(portfolio[2,2][[1]])
```
<aside>Note: Each element of the matrix is wrapped as a list so that they are of equal length. To access the underlying object, the ichimoku cloud in this case, we simply extract it using [[1]].</aside>

^[Further examples: Youngju Nielsen of Sungkyunkwan University uses do.call / lapply to good effect in her course https://www.coursera.org/learn/the-fundamental-of-data-driven-investment/]

