---
title: "Efficient R: do.call / lapply"
description: |
  A distinctive coding style
author:
  - name: shikokuchuo
date: 05-28-2021
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
par(mar = c(0, 0, 0, 0))
plot(imager::load.image(photos), axes = FALSE)

```
<aside> Shikokuchuo </aside>

### Use case

The use of the do.call / lapply() combination is a powerful way to leverage functional programming in R. In short, write a function that performs some actions and apply it to a list of inputs, which is then combined into a single object.

Let us take an example, where we would like to calculate the ichimoku clouds for a selection of the major world stock indices, but also preserve the volume data, all in one tidy object.

We use the 'ichimoku' package^[Gao, C. (2021), *ichimoku: Visualization and Tools for Ichimoku Kinko Hyo Strategies*. R package version 1.1.0, <https://CRAN.R-project.org/package=ichimoku>. Note that since version 1.2.0 additional data such as volume may be retained within the ichimoku object itself.] which not only draws the ichimoku clouds, but also provides an interface to the OANDA fxTrade API which is a rich source of high-quality financial data (free but requires registration).

### Setup

We could set it up as per the below:

- tickers: a vector defining the stock symbols in our portfolio
- process(): a function that generates a row in a data frame or matrix

```{r functional, echo=TRUE}
library(ichimoku)

tickers <- c("DE30_EUR", "JP225_USD", "SPX500_USD", "UK100_GBP")

process <- function(x, from, to) {
  # Use ichimoku::oanda() to retrieve data from the OANDA fxTrade API
  pxdata <- oanda(x, from = from, to = to)
  # Extract volume column
  volume <- pxdata$volume
  # Calculate the cloud by calling ichimoku::ichimoku()
  cloud <- ichimoku(pxdata, ticker = x)
  # Return a list of ticker, ichimoku cloud object, volume data
  list(x, cloud, volume)
}

```
<aside> Note: the original pricing data is preserved within the ichimoku object.</aside> 

We now want to apply our function to each element of 'tickers' in turn, and then for the results to be combined.

### Loop

One way to achieve this would be to iterate over 'tickers' using a loop:

```{r loop, echo=TRUE}

# Define a list to contain the loop output, specifying the length in advance as good practice
portfolio <- vector(mode = "list", length = length(tickers))

# Loop over each element in 'tickers' and save in pre-defined list
for (i in seq_along(tickers)) {
  portfolio[[i]] <- process(tickers[i], from = "2015-09-03", to = "2016-06-30")
}

# Create output matrix by calling rbind on each element of the list
portfolio <- do.call(rbind, portfolio)

portfolio

```

This approach takes 3-4 lines of code.

Furthermore, 'i' remains as a leftover object in the global environment.

Somewhat messy.

### do.call / lapply

Instead we can use a do.call / lapply() combination to achieve the same result in one line:

```{r lapply, echo=TRUE}

portfolio <- do.call(rbind, lapply(tickers, process, from = "2015-09-03", to = "2016-06-30"))

portfolio

```

There are also no intermediate objects generated that clutter the global environment.

To explain:

- First lapply() applies to a list or list-like object ('tickers'), a function ('process'). The arguments to the function are supplied immediately afterwards. lapply always returns a list.

- This is then fed into do.call(), which calls a function ('rbind') on a list of arguments (the output of 'lapply', a list). This creates a matrix.

The use of do.call / lapply() provides for a far more succinct and distinctive coding style.

The added bonus is that of the 'apply' family of functions, lapply() is almost always the fastest and most performant as the output type is fixed and it does not try to do things with names or simplify the output structure.

For a more structured format than a list, lapply() can be fed into a do.call() with:

- c() to form a vector
- cbind() or rbind() to form a matrix

The use of this type of combination is of particular benefit in programming where both performance and predictability of output types is paramount.

### Tidy data output

```{r portfolio, echo=TRUE}
portfolio
```

'portfolio' is a tidy matrix with a row for each ticker, and a column for each data type.

We can easily access any element of the matrix by specifying its index value, for example the ichimoku object for the S&P 500 Index by [3,2]. In the example below we run autostrat() on this object:

```{r output, echo=TRUE}
autostrat(portfolio[3, 2][[1]], n = 1)
```
<aside>Note: Each element of the matrix is wrapped as a list so that they are of equal length. To access the underlying object, the ichimoku object in this case, we simply extract it using [[1]].</aside>

^[Further examples: Youngju Nielsen of Sungkyunkwan University uses do.call / lapply to good effect in her course <https://www.coursera.org/learn/the-fundamental-of-data-driven-investment/>]

