---
title: "Re-introducing mirai - a minimalist async evaluation framework for R"
description: |
  Concurrent code execution with maximum flexibility and automatic resolution
author:
  - name: shikokuchuo
date: 2022-04-18
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
par(mar = c(0, 0, 0, 0))
plot(imager::load.image(photos), axes = FALSE)

```
<aside>Shikokuchuo</aside>
<br />
{mirai} is a minimalist async evaluation framework for R.

未来 みらい mirai is Japanese for 'future'.

[mirai](https://cran.r-project.org/package=mirai) provides an extremely simple and lightweight method for concurrent / parallel code execution.

Since the original ['introduction' article](https://shikokuchuo.net/posts/16-introducing-mirai/) two months ago, mirai is currently in its fourth incarnation, hence the need for a re-introduction. In this time, the package has also featured in RStudio's [Top 40 New CRAN Packages for February 2022](https://rviews.rstudio.com/2022/03/28/february-2022-top-40-new-cran-packages/).

We outline some of the main visible innovations below, but suffice to say that the backend (especially the code behind the [nanonext](https://cran.r-project.org/package=nanonext) package) has been very much optimised over the course of this time as well.

### 1. Two Functions -> One Function

Originally, the package revolved around 2 functions - `eval_mirai()` and `call_mirai()`. Those are still there, and work the same way - however the package has managed to become even more minimalist with only one being required now.

This one function is simply called `mirai()`.

(`eval_mirai()` maps to it as an alias.)

Using `mirai()` returns a ‘mirai’ object immediately.

A mirai evaluates an arbitrary expression asynchronously, resolving automatically upon completion.

### 2. Automatic Resolution

Yes, resolving automatically, hence `call_mirai()` is no longer needed (although it can still sometimes be useful to call and wait for results).

The evaluated result of a mirai is stored at `$data`. If the asynchronous operation has yet to complete then this will return an 'unresolved' logical NA value. That is an actual NA value of type logical classed as 'unresolved'.

As soon as it resolves, then `$data` will return the actual value of the evaluated expression.

Enter `unresolved()`, a helper function that can be used in control flow statements to allow you to do things before and after resolution of a mirai. This means that you never have to actually just *wait* on a mirai.

The code output below probably serves as a better demonstration than a lengthy explanation:

#### Example using a while loop:

```{r unresolved}
library(mirai)

m <- mirai({is.null(Sys.sleep(n)) && return(TRUE)}, n = 1)
m$data
while (unresolved(m)) {
  cat("unresolved\n")
  Sys.sleep(1)
} 
m$data

```

#### Equivalent using a repeat loop:

```{r unresolved2}

m <- mirai({is.null(Sys.sleep(n)) && return(TRUE)}, n = 1)
m$data
repeat {
  unresolved(m) || break
  cat("unresolved\n")
  Sys.sleep(1)
} 
m$data

```

### 3. Arbitrary expressions

You may have noticed in the above examples that the expression being evaluated was wrapped in `{ }`. mirai supports this being an arbitrarily-long, possibly multi-line expression. Similarly there is no limit to the number of arguments supplied to the call.

To supply objects that are already present in the calling environment, these may simply be passed in as per the below:


```{r passing in variables}
mat <- matrix(c(1, 2, 3, 4), ncol = 2)
mat

m <- mirai({
  x <- t(x)
  as.data.frame(x)
}, x = mat)

call_mirai(m)$data

```

### 4. Daemons

The default behaviour is to just spin up background processes as required. This offers maximum simplicity and ease of use - no need to consider the backend at all.

First of all this is remarkable in itself, in that mirai simply works on all platforms R can be installed - across Linux, Windows, Mac, Solaris etc. and this is down to the cross-platform support built in to the underlying NNG C library that [nanonext](https://cran.r-project.org/package=nanonext) provides a binding for.

Second, the startup time is relatively tiny as a completely clean background R process with '--vanilla' settings is started each time. However for the highest performance applications, there is now the option to start up a set number of processes (daemons) upfront, and these will be ready to wait for instructions, achieveing even more minimal latency.

Setting daemons is as simple as:

```{r daemons, eval=FALSE}
daemons(8)
#> [1] 8
```
.. and 8 daemons are created.

### Wrap-up

We presented above 4 key updates for the fourth version of [mirai](https://cran.r-project.org/package=mirai) (v0.4.0). We invite you to try it out for yourselves:

```{r install, eval=FALSE}
install.packages("mirai")
```

Any issues/comments please feed back at Github: <https://github.com/shikokuchuo/mirai>

Package website: <https://shikokuchuo.net/mirai/>
