---
title: "nanonext - High Performance Persistent HTTP Sessions"
description: |
  ncurl sessions
author:
  - name: shikokuchuo
date: 2023-01-26
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
par(mar = c(0, 0, 0, 0))
plot(imager::load.image(photos), axes = FALSE)

```
<aside>Shikokuchuo</aside>
<br />
Persistent http(s) sessions is a new feature added in nanonext 0.7.3.

This allows for efficient polling by keeping an open connection with the server, transacting as and when needed. 

Provides an ideal, low-latency solution to requesting real time data over a REST API, especially when there are limits in place for the frequency of new connections.

### ncurl_session()

Create a session (persistent connection):

```{r ncurl rest}

library(nanonext)
sess <- ncurl_session("https://httpbin.org/headers")

```

### transact()

Transact over the session (repeatedly if required):

```{r ncurl result}

res <- transact(sess)
res

```

### Timing

Allows much lower latencies in returning results:

```{r ncurl timing}

library(microbenchmark)

microbenchmark(transact(sess), ncurl("https://httpbin.org/headers"))

```

### Links

{nanonext} package website: <https://shikokuchuo.net/nanonext/>
<br />On CRAN: <https://cran.r-project.org/package=nanonext>

{nanonext} features in the 'Web Technologies' CRAN Task View under 'Core Tools For HTTP Requests': <https://cran.r-project.org/view=WebTechnologies>
and also in the 'High Performance Computing' CRAN Task View:
<https://cran.r-project.org/view=HighPerformanceComputing>

