---
title: "Resources for global sustainability"
description: |
  A compendium
author:
  - name: shikokuchuo
date: 05-02-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
develop <- function(x) {
  plot(imager::load.image(x), axes = FALSE)
  paste0(openssl::sha256(file(x)))
}
par(mar = c(0, 0, 0, 0))
data.frame(sha256 = develop(photos))

```
<aside> Dawn over Shikokuchuo </aside>

### Stockholm Resilience Center

Sustainability science for biosphere stewardship, Stockholm University

Video: Our future in the Anthropocene biosphere

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/5pAONwIKor8" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### Svalbard Global Seed Vault

Safeguarding Seeds for the Future

Established and funded by the Norwegian Ministry of Agriculture and Food.

Provides safe, free and long-term storage of seed duplicates from all genebanks and nations participating in the global community’s joint effort to ensure the world’s future food supply. 

The facility serves a humanitarian purpose and is part of the international system for conserving plant genetic diversity guided by the UN organisation for Food and Agriculture (FAO).

https://www.seedvault.no/

### The National Academies of Sciences, Engineering, and Medicine

More than 100 leading scientists including many Nobel Prize winners issued a statement following two days of scientific deliberations at the first Nobel Prize Summit, 26-28 April 2021.

Excerpts from the statement:

```{r statement, echo=TRUE, eval=FALSE}
Global heating and habitat loss amount to nothing less than a vast and
uncontrolled experiment on Earth’s life-support system.

Time is running out to prevent irreversible changes.

The remaining carbon budget for a 67% probability of not exceeding 1.5°C global
warming will be exhausted before 2030. 

Global sustainability offers the only viable path to human safety, equity,
health, and progress.

Humanity is waking up late to the challenges and opportunities of active
planetary stewardship. But we *are* waking up.

```

Full text: <br />
https://www.nationalacademies.org/news/2021/04/nobel-prize-laureates-and-other-experts-issue-urgent-call-for-action-after-our-planet-our-future-summit

### Openlearn, The Open University

Greta Thunberg: A Year to change the World

Encounters with some of the world’s leading scientists and economists allow the series to examine what the latest science tells us about what can be done to avert the worst effects of climate change.

https://www.open.edu/openlearn/tv-radio-events/tv/greta-thunberg-year-change-the-world

### TED

Stephen Hawking <br />
*Questioning the universe* <br />
TED February 2008

<div style="max-width:854px"><div style="position:relative;height:0;padding-bottom:56.25%"><iframe src="https://embed.ted.com/talks/lang/en/stephen_hawking_questioning_the_universe" width="854" height="480" style="position:absolute;left:0;top:0;width:100%;height:100%" frameborder="0" scrolling="no" allowfullscreen></iframe></div></div>

<aside>
Transcript: 09:06 | 
Stephen Hawking: I think it quite likely that we are the only civilization within several hundred light years; otherwise we would have heard radio waves. **The alternative is that civilizations don't last very long, but destroy themselves**.
</aside>

### UI design x data science x climate change

What can a technologist do about climate change?

Bret Victor: A personal view.

http://worrydream.com/ClimateChange/
