---
title: "Github Actions with R"
description: |
  Deploy and automate your code to the cloud
author:
  - name: shikokuchuo
date: 05-04-2021
categories:
  - R
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

```{r index, echo=FALSE}

photos <- file.path("_images", list.files("_images"))
develop <- function(x) {
  plot(imager::load.image(x), axes = FALSE)
  paste0(openssl::sha256(file(x)))
}
par(mar = c(0, 0, 0, 0))
data.frame(sha256 = develop(photos))

```
<aside>Shikokuchuo</aside>

### Github Actions with R

Set up a cron job to run your R scripts at specified times.

Enabled with a simple yaml configuration file.

Save the following as main.yml in .github/workflows of your Github repository:

```{r yml, eval=FALSE}
name: Action
on:
  schedule:
    - cron: '30 22 * * 1-5'
jobs:
  render:
    runs-on: windows-latest
    steps:
      - uses: actions/checkout@v2
      - name: Set up R
        uses: r-lib/actions/setup-r@v1
      - name: Install dependencies
        run: |
          install.packages(c("magrittr", "purrr"), type = "win.binary")
        shell: Rscript {0}
      - name: Script
        run: Rscript nameofscript.R
      - name: Commit files
        run: |
          git config --local user.name github-actions
          git config --local user.email "actions@github.com"
          git add docs/*
          git commit -am "commit on $(date)"
          git push origin main
        env:
          REPO_KEY: ${{secrets.GITHUB_TOKEN}}
          username: github-actions
```
<aside>
1. This example cron job runs every Mon-Fri at 22.30.

2. Customize your R packages to install.

3. Make sure to change nameofscript.R to your actual script name.

4. Assumes your script writes files to the 'docs' directory, change if necessary.

5. Assumes your repository branch is 'main', change if necessary.

6. Note that this script is run on a Windows VM using Windows R binary packages. This is currently much faster than building a lot of dependencies on Linux (which is also prone to failure).
</aside>

To generate R Markdown documents (or for that matter render a Distill website), you will want to add the following step after 'Set up R':

```{r yml2, eval=FALSE}
      - name: Set up Pandoc
        uses: r-lib/actions/setup-pandoc@v1
```

### Reference

The Github Actions with R reference: <br />
https://orchid00.github.io/actions_sandbox/
